---
title: "Evaluating Pennsylvania Re-employment Bonus Experiment"
author: "Jerry Hong"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

```{r load libraries, include=FALSE}
pacman::p_load(tidyverse, estimatr, knitr, tinytex)
```

## Background

The Pennsylvania Re-employment Bonus Experiment was a social experiment conducted in the 1980s to evaluate the effectiveness of a re-employment bonus program. The program was designed to encourage unemployed workers to find new jobs by offering a cash bonus to those who found new jobs within a certain time frame. These experiments were conducted to test the incentive effects of alternative compensation schemes for unemployment insurance (UI). UI claimants are randomly assigned to a control group or one of five treatment groups. The control group is eligible for regular UI benefits, while the treatment groups are eligible for regular UI benefits plus a bonus if they find a job within a certain time frame, provided that the job was retained for a specific duration. This study will focus on the comparison of the control group and merged treatment groups 4 and 6 and analyze the treatment effects of these incentives.

```{r load data, include=FALSE}
penn <- read.table("https://psantanna.com/Econ520/files/penn_jae.dat", header=TRUE)
```

## Data Exploration and Cleaning

The data set has 13,913 observations across 23 variables. However, we are subsetting this with the control and merged treatment groups. We will merge treatment groups 4 and 6 into one group and subset the data including the control group 0 and treatment group 4. 

```{r merge}
penn_subset <- penn %>%
  mutate(tg = ifelse(tg %in% c(4, 6),4,tg)) %>%
  filter(tg %in% c(0, 4))
```

With this subset, we have 6,384 observations with 3,354 in the control and 3,030 in the treatment group. There are also no missing data in the subset.


## Analyze Treatment Effects

## Takeaways

## Next steps

## Limitations


