---
title: "Evaluating Pennsylvania Re-employment Bonus Experiment"
author: "Jerry Hong"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

```{r load libraries, include=FALSE}
pacman::p_load(tidyverse, estimatr, knitr, tinytex)
```

## Background

The Pennsylvania Re-employment Bonus Experiment was a social experiment conducted in the 1980s to evaluate the effectiveness of a re-employment bonus program. The program was designed to encourage unemployed workers to find new jobs by offering a cash bonus to those who found new jobs within a certain time frame. These experiments were conducted to test the incentive effects of alternative compensation schemes for unemployment insurance (UI). UI claimants are randomly assigned to a control group or one of five treatment groups. The control group is eligible for regular UI benefits, while the treatment groups are eligible for regular UI benefits plus a bonus if they find a job within a certain time frame, provided that the job was retained for a specific duration. This study will focus on the comparison of the control group and merged treatment groups 4 and 6 and analyze the treatment effects of these incentives.

```{r load data}
penn <- read.table("https://psantanna.com/Econ520/files/penn_jae.dat", header=TRUE)
```

## Data Exploration and Cleaning

The data set has 13,913 observations across 23 variables. However, we are subsetting this with the control and merged treatment groups. We will merge treatment groups 4 and 6 into one group and subset the data including the control group 0 and treatment group 4. 

```{r merge, include=FALSE}
penn_subset <- penn %>%
  mutate(tg = ifelse(tg %in% c(4, 6),4,tg)) %>%
  filter(tg %in% c(0, 4))
```

With this subset, we have 6,384 observations with 3,354 in the control and 3,030 in the treatment group. There are also no missing data in the subset.

```{r summary stats, echo=FALSE}
penn_subset %>%
  mutate(log_duration = log(inuidur1)) %>%
  group_by(tg) %>%
  summarize(count=n(),
            avg_unemp_duration = mean(inuidur1, na.rm = TRUE),
            avg_log_duration = mean(log_duration, na.rm = TRUE),
            median_unemp_duration = median(inuidur1, na.rm = TRUE),
            median_log_duration = median(log_duration, na.rm = TRUE),
            sd_unemp_duration = sd(inuidur1, na.rm = TRUE),
            sd_log_duration = sd(log_duration, na.rm = TRUE))
```

Between the control and treatment groups, the difference in the unemployment duration, even when taking the log, is not as substantial. The control group has a slight increase in duration with roughly 0.9 more weeks of unemployment. Both the median and standard deviations are also marginally similar, indicating slight variations in the distribution of unemployment duration.

Below shows the proportions of the control and treatment groups based on gender, race, and age.

```{r proprotions, echo=FALSE}
# prop table for female
prop_female <- with(penn_subset, prop.table(table(tg, female), margin =2))

# prop table for black
prop_black <- with(penn_subset, prop.table(table(tg, black), margin =2))

# prop table for othrace
prop_othrace <- with(penn_subset, prop.table(table(tg, othrace), margin =2))

# prop table for agelt35
prop_agelt35 <- with(penn_subset, prop.table(table(tg, agelt35), margin =2))

# prop table for agegt54
prop_agegt54 <- with(penn_subset, prop.table(table(tg, agegt54), margin =2))

prop_df <- cbind(prop_female, prop_black, prop_othrace, prop_agelt35, prop_agegt54)

colnames(prop_df) <- c("Male", "Female",
                                    "Not Black", "Black",
                                    "Not Other Race", "Other Race",
                                    "Not Younger 35", "Younger 35",
                                    "Not Older 54", "Older 54")
row.names(prop_df) <- c("Control", "Treatment")
prop_df
```

The proportions in the control group are generally higher than the treatment group across all categories, except those who are of other race. 

## Analyze Treatment Effects

## Takeaways

## Limitations

There are a couple of limitations to this study. One of which is that the data is from the 1980s, which may not be as relevant to today's labor market with new policies enacted since that study. Additionally, the data is from Pennsylvania, which may not be generalized to other states or regions as they may have different labor regulations. Lastly, the study only focuses on the comparison of the control group and merged treatment groups 4 and 6, which may not capture the full effects of the re-employment bonus program.

## Next steps

Some next steps to take with this study include: a more general study encompassing more states and regions besides Pennsylvania. Additionally, a more comprehensive analysis of all treatment groups to understand the full effects of the re-employment bonus program. Lastly, a more updated study to reflect the current labor market and policies in place.
